---
title: "The Law of Large Numbers: Single Die Roll"
format: html
execute:
  warning: false
  message: false
---

## 1. RESET AND SIMULATE

### (Event: Rolling a 1 or a 5)

This cell clears memory and simulates rolling a single die. We are tracking how often we get a **1** or a **5**.

```{r}
#| label: simulation-setup

# THE AUTOMATIC BROOM
rm(list = ls()) 

# Number of total trials
N <- 10000

# Generate N die rolls
die_rolls <- sample(1:6, N, replace = TRUE)

# Logical check: Is the roll a 1 or a 5?
is_target <- die_rolls %in% c(1, 5)

# Combine into data frame
experiment_data <- data.frame(
  Trial = 1:N,
  Die_Rolled = die_rolls,
  Success = is_target
)

# Calculate running probability (Relative Frequency)
running_probability <- cumsum(is_target) / 1:N
```

## 2. Manual Verification: The Data Table

Adjust `n_show` to show the students the raw results. They can count the number of TRUEs and divide by the Trial number.

```{r}
#| label: manual-table
n_show <- 10 

head(experiment_data, n_show)
```

## 3. Snapshot Analysis: Computing Specific Trial Slices

This new section allows you to choose a specific number of early trials to see the calculated probability at that exact moment.

```{r}
#| label: snapshot-analysis

# CHOOSE YOUR NUMBER: Set how many first trials to analyze (e.g., 25, 50, 100)
n_slice <- 10000

# Calculate successes in this specific slice
slice_successes <- sum(experiment_data$Success[1:n_slice])
slice_fraction <- slice_successes / n_slice

# Output the results directly into the report
cat("--- Analysis for the First", n_slice, "Trials ---\n")
cat("Total Successes (1 or 5): ", slice_successes, "\n")
cat("Relative Frequency      : ", round(slice_fraction, 4), "\n")
cat("Theoretical Target      : ", round(1/3, 4), "\n")
```

## 4. Manual Verification: The Plot (Early Trials)

This shows the "chaos" at the start of the experiment.

```{r}
#| label: manual-plot
#| fig-width: 8
#| fig-height: 4

n_show_plot <- 10 

plot(1:n_show_plot, running_probability[1:n_show_plot], 
     type = "b", pch = 19, col = "blue", lwd = 2,
     ylim = c(0, 1), xlab = "Trial Number", ylab = "Running Frequency",
     main = paste("First", n_show_plot, "Trials"))

# Theoretical probability line (2/6 = 1/3)
abline(h = 1/3, col = "red", lwd = 2, lty = 2)
```

## 5. The Law of Large Numbers (Convergence)

As the trials increase, the "noise" disappears and the frequency stabilizes.

```{r}
#| label: full-convergence
#| fig-width: 8
#| fig-height: 5

n_limit <- N

plot(1:n_limit, running_probability[1:n_limit], 
     type = "l", col = "blue", ylim = c(0, 1), 
     xlab = "Number of Trials", ylab = "Running Frequency",
     main = paste("Convergence over", n_limit, "Trials"))

abline(h = 1/3, col = "red", lwd = 2, lty = 2)

legend("topright", legend = c("Running Frequency", "Theoretical (1/3)"), 
       col = c("blue", "red"), lty = c(1, 2), lwd = c(1.5, 2))
```

## 6. Analyzing the Error (Convergence Speed)

The log-log plot shows the error magnitude decreasing.

```{r}
#| label: error-analysis
#| fig-width: 8
#| fig-height: 5

n_limit_err <- N

# Calculate absolute difference from 1/3
raw_error <- abs(running_probability[1:n_limit_err] - (1/3))

# Floor the zero-errors to 1e-6 for the log plot spikes
visual_error <- ifelse(raw_error == 0, 1e-6, raw_error)

plot(1:n_limit_err, visual_error, 
     type = "l", log = "xy", col = "darkgreen",
     yaxt = "n", 
     xlab = "Number of Trials (Log Scale)", 
     ylab = "Absolute Error (Log Scale)",
     main = "Error Decay: Convergence towards Zero")

# Custom Y-axis labels
axis(2, at = c(1, 1e-1, 1e-2, 1e-3, 1e-6), 
     labels = c("1", "0.1", "0.01", "0.001", "PERFECT"), 
     las = 1)

grid(equilogs = TRUE, col = "lightgray", lty = "dotted")
abline(h = 1e-6, col = "gray", lty = 3)
```
