---
title: "The Law of Large Numbers: A Two-Stage Experiment"
format: html
execute:
  warning: false
  message: false
---

## 1. RESET AND SIMULATE

### (Click the Green Triangle to Clear Memory and Run Fresh)

This stage involves a conditional experiment: We roll a die; if it's a 1 or 2, the experiment stops. If it's a 3, 4, 5, or 6, we flip a coin. We are tracking the probability of landing on **Tails (T)**.

```{r}
#| label: simulation-setup

# THE AUTOMATIC BROOM: This clears all variables from R's memory.
rm(list = ls()) 

# Number of total trials
N <- 1000000

# Generate data
die_rolls <- sample(1:6, N, replace = TRUE)
coin_flips <- sample(c("H", "T"), N, replace = TRUE)

# Logic: If die is 1 or 2 -> "Stop". Otherwise, take the coin flip result.
final_outcomes <- ifelse(die_rolls %in% c(1, 2), "Stop", coin_flips)

experiment_data <- data.frame(
  Trial = 1:N,
  Die_Rolled = die_rolls,
  Final_Outcome = final_outcomes
)

# Calculate running probability for Tails (T)
is_tails <- experiment_data$Final_Outcome == "T"
running_probability <- cumsum(is_tails) / 1:N
```

## 2. Manual Verification: The Data Table

Adjust `n_show` to see the individual results of the die rolls and the resulting outcomes.

```{r}
#| label: manual-table
# CHANGE THIS NUMBER to show more or fewer rows:
n_show <- 10 

head(experiment_data, n_show)
```

## 3. Snapshot Analysis: Computing Specific Trial Slices

This section allows you to pick a specific "cut-off" point early in the experiment to see what the relative frequency looks like at that exact moment.

```{r}
#| label: snapshot-analysis

# CHOOSE YOUR NUMBER: Set how many first trials to analyze (e.g., 10, 25, 50, 100)
n_slice <- 400

# Extract the subset and count how many times 'T' appeared
subset_results <- experiment_data$Final_Outcome[1:n_slice]
tails_count <- sum(subset_results == "T")

# Calculate the relative frequency for this slice
slice_freq <- tails_count / n_slice

# Output the findings
cat("--- Snapshot Analysis for First", n_slice, "Trials ---\n")
cat("Total Tails Observed :", tails_count, "\n")
cat("Relative Frequency   :", round(slice_freq, 4), "\n")
cat("Theoretical Target   :", round(1/3, 4), "\n")
```

## 4. Manual Verification: The Plot (Early Trials)

The early trials usually show high volatility. As you change `n_show` here, you can see how much the "luck" of the first few flips swings the graph.

```{r}
#| label: manual-plot
#| fig-width: 8
#| fig-height: 4

# CHANGE THIS NUMBER to match the table or show a specific range:
n_show_plot <- 50

plot(1:n_show_plot, running_probability[1:n_show_plot], 
     type = "b", pch = 19, col = "blue", lwd = 2,
     ylim = c(0, 1), xlab = "Trial Number", ylab = "Running Frequency",
     main = paste("First", n_show_plot, "Trials"))

# Theoretical line: P(Not Stop) * P(Tails) = (4/6) * (1/2) = 1/3
abline(h = 1/3, col = "red", lwd = 2, lty = 2)
```

## 5. The Law of Large Numbers (Convergence)

This view shows the full scale of the experiment. Notice how the blue line eventually "sticks" to the red theoretical line.

```{r}
#| label: full-convergence
#| fig-width: 8
#| fig-height: 5

n_limit <- N 

plot(1:n_limit, running_probability[1:n_limit], 
     type = "l", col = "blue", ylim = c(0, 1), 
     xlab = "Number of Trials", ylab = "Running Frequency",
     main = paste("Convergence over", n_limit, "Trials"))

abline(h = 1/3, col = "red", lwd = 2, lty = 2)

legend("topright", legend = c("Running Frequency", "Theoretical (1/3)"), 
       col = c("blue", "red"), lty = c(1, 2), lwd = c(1.5, 2))
```

## 6. Analyzing the Error (Convergence Speed)

This log-log plot visualizes how the "distance" between our experimental results and the truth (1/3) shrinks as we collect more data.

```{r}
#| label: error-analysis
#| fig-width: 8
#| fig-height: 5

n_limit_err <- N

# Calculate error
raw_error <- abs(running_probability[1:n_limit_err] - (1/3))

# Floor the zero-errors to 1e-6 for the log scale
visual_error <- ifelse(raw_error == 0, 1e-6, raw_error)

plot(1:n_limit_err, visual_error, 
     type = "l", log = "xy", col = "darkgreen",
     yaxt = "n", 
     xlab = "Number of Trials (Log Scale)", 
     ylab = "Absolute Error (Log Scale)",
     main = "Error Decay: Convergence towards Zero")

# Custom Y-axis labels
axis(2, at = c(1, 1e-1, 1e-2, 1e-3, 1e-6), 
     labels = c("1", "0.1", "0.01", "0.001", "PERFECT"), 
     las = 1)

grid(equilogs = TRUE, col = "lightgray", lty = "dotted")
abline(h = 1e-6, col = "gray", lty = 3)
```
