---
title: "Coin Flip Simulation"
format: html
editor: source
---

## Simulation Setup

This document performs a coin flip simulation. We begin by loading the necessary visualization library and defining our core parameters: the total number of simulations and the theoretical probability of flipping heads.

```{r}
#| label: setup
#| message: false
#| warning: false

# THE AUTOMATIC BROOM
rm(list = ls()) 

# Load ggplot2 for visualization
library(ggplot2)

# Define simulation parameters
total_sims <- 1000
prob_heads <- 0.05

# Note: set.seed() is omitted to ensure non-repeatable, random results on each run.
```

## Executing the Simulation

Next, we run the simulation using the `sample` function. The results are stored in a data frame so they can be easily plotted and analyzed in the subsequent sections.

```{r}
#| label: simulate

# Simulate the coin flips
flips <- sample(
  x = c("Heads", "Tails"),
  size = total_sims,
  replace = TRUE,
  prob = c(prob_heads, 1 - prob_heads)
)

# Convert to data frame
results_df <- data.frame(Outcome = flips)
```

## Visualizing the Results

The following bar plot visualizes the frequency of Heads versus Tails across all simulated trials.

```{r}
#| label: visualization
#| echo: false

# Create a bar plot of the outcomes
ggplot(results_df, aes(x = Outcome)) +
  geom_bar(fill = c("steelblue", "darkgray"), color = "black", width = 0.6) +
  theme_minimal() +
  labs(
    title = paste("Histogram of", total_sims, "Coin Flips"),
    x = "Coin Face",
    y = "Count"
  )
```

## Summary Statistics

Finally, we calculate the exact totals and fractions for each outcome to see how closely our empirical results match the theoretical probability.

```{r}
#| label: statistics
#| echo: false

# Calculate total occurrences
heads_count <- sum(results_df$Outcome == "Heads")
tails_count <- sum(results_df$Outcome == "Tails")

# Calculate fractions
heads_fraction <- heads_count / total_sims
tails_fraction <- tails_count / total_sims

# Output the final statistics
cat("--- Simulation Results ---\n")
cat("Total Simulations Ran :", total_sims, "\n")
cat("Target Prob. of Heads :", prob_heads, "\n\n")

cat("Total Heads           :", heads_count, "\n")
cat("Total Tails           :", tails_count, "\n\n")

cat("Fraction of Heads     :", heads_fraction, "\n")
cat("Fraction of Tails     :", tails_fraction, "\n")
```
