---
title: "The Central Limit Theorem: Parameter Breakdown"
author: "Probability Class Demonstration"
format: html
---

## 🎯 The Central Limit Theorem (CLT)

The Central Limit Theorem states that the distribution of sample means approximates a normal distribution as the sample size $n$ increases.

$$ \bar{X} \sim N\left(\mu, \frac{\sigma^2}{n}\right). $$

```{r}
#| label: setup
#| message: false
#| warning: false

library(ggplot2)
set.seed(42)
```

------------------------------------------------------------------------

## 🧠 Clarifying the Three Parameters

To simulate the CLT, we strictly require three distinct numbers. Think of this as drawing balls from a giant bucket.

-   🌍 **1. `universe_size`:** The total number of balls inside the giant bucket. This defines our underlying, non-normal population. We only generate this bucket once.
-   🥄 **2. `sample_size` (**$n$): The size of our scoop. How many balls do we pull out of the bucket *at one time* to calculate a single average?
-   🔄 **3. `repetitions`:** How many times do we dip our scoop into the bucket, calculate the average, and write that average down on a piece of paper? We need a large number of repetitions to draw a smooth bell curve at the end.

------------------------------------------------------------------------

## ⚙️ 1. Parameter Configuration

Define the mathematical environment using the three distinct parameters explained above.

```{r}
#| label: parameters

# 📊 DISTRIBUTION TYPE
# Choose the shape of the balls in the bucket: "Exponential", "Uniform", or "U-Shaped"
chosen_distribution <- "Exponential"

# 🌍 PARAMETER 1: THE BUCKET
# How many total items exist in our underlying universe?
universe_size <- 1000000

# 🔄 PARAMETER 3: THE REPETITIONS
# How many sample means will we calculate to build the final normal curve?
repetitions <- 50000
```

------------------------------------------------------------------------

## 📉 2. Generating the Universe

We fill the bucket based on your chosen distribution.

```{r}
#| label: define-population

# Fill the bucket with 'universe_size' number of items
if (chosen_distribution == "Exponential") {
  bucket_of_data <- rexp(universe_size, rate = 1)
} else if (chosen_distribution == "Uniform") {
  bucket_of_data <- runif(universe_size, min = 0, max = 10)
} else if (chosen_distribution == "U-Shaped") {
  bucket_of_data <- rbeta(universe_size, shape1 = 0.5, shape2 = 0.5)
} else {
  stop("Invalid distribution selected.")
}

pop_data <- data.frame(Value = bucket_of_data)

# 🖼️ Plot the entire bucket to show its non-normal shape
# boundary = 0 strictly aligns the first bin to prevent negative space leaking
ggplot(pop_data, aes(x = Value)) +
  geom_histogram(bins = 50, fill = "gray30", color = "white", boundary = 0) +
  labs(
    title = paste("The Underlying Universe:", chosen_distribution), 
    subtitle = paste("Total items in bucket:", format(universe_size, scientific = FALSE)),
    y = "Frequency"
  ) +
  theme_minimal()
```

------------------------------------------------------------------------

## 🎲 3. The Sampling Function

This function executes the "scooping" process. It requires the bucket, the size of the scoop (`sample_size`), and the number of times to dip the scoop (`repetitions`).

```{r}
#| label: sampling-function

get_sample_means <- function(bucket, scoop_size_n, total_dips) {
  
  # Create an empty list to store the averages we calculate
  recorded_averages <- numeric(total_dips)
  
  # Dip into the bucket 'total_dips' times
  for (i in 1:total_dips) {
    
    # Scoop out 'scoop_size_n' items
    current_scoop <- sample(bucket, size = scoop_size_n, replace = TRUE)
    
    # Calculate the average of this single scoop and record it
    recorded_averages[i] <- mean(current_scoop)
  }
  
  return(data.frame(
    Sample_Mean = recorded_averages, 
    Sample_Size_Label = paste("n =", scoop_size_n)
  ))
}
```

------------------------------------------------------------------------

## 🚀 4. Executing the Simulation

We test the theory by changing **Parameter 2** (`sample_size` or $n$). We will use a scoop of 1, a scoop of 10, and a scoop of 30. We repeat each scooping process 50,000 times.

```{r}
#| label: run-simulation

# 🥄 PARAMETER 2: THE SCOOP SIZE (n)
# Execute the simulation for n=1, n=10, and n=30
results_n1  <- get_sample_means(bucket_of_data, scoop_size_n = 1,  total_dips = repetitions)
results_n10 <- get_sample_means(bucket_of_data, scoop_size_n = 10, total_dips = repetitions)
results_n30 <- get_sample_means(bucket_of_data, scoop_size_n = 3000, total_dips = repetitions)

# Combine the recorded averages for plotting
all_results <- rbind(results_n1, results_n10, results_n30)

# Lock factor levels to maintain sequential plotting
all_results$Sample_Size_Label <- factor(
  all_results$Sample_Size_Label, 
  levels = unique(all_results$Sample_Size_Label)
)

# 🖼️ Plot the progression
# Added boundary = 0 to prevent spurious non-monotonicity at n = 1
ggplot(all_results, aes(x = Sample_Mean)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, fill = "steelblue", color = "white", boundary = 0) +
  facet_wrap(~Sample_Size_Label, scales = "free") +
  labs(
    title = paste("CLT Convergence:", chosen_distribution, "Universe"), 
    subtitle = "Notice the progression toward a symmetrical normal distribution as the scoop size (n) increases."
  ) +
  theme_minimal()
```
