---
title: "First Steps in R and Data Visualization"
---

## 1. Welcome to RStudio and Quarto (5 Minutes)

Welcome to your first session. R is the programming language, and RStudio is the integrated development environment (IDE)—the software interface we use to interact with R.

RStudio is divided into four main panes: \* **Source (Top Left):** This is the text editor where we write reproducible documents like this Quarto file. \* **Console (Bottom Left):** The engine. When code runs, it executes here. \* **Environment (Top Right):** The workspace memory. It lists all the data and variables R is currently storing. \* **Files/Plots/Help (Bottom Right):** A utility pane to navigate your computer, view generated charts, and read the official R documentation.

### Understanding This Document

This is a **Quarto Document**. It allows us to combine human-readable text (what you are reading now) with executable computer code. Code is contained inside gray boxes called **chunks**.

To run a chunk of code, click the green "Play" button in the top-right corner of the chunk, or place your cursor on a specific line and press `Ctrl + Enter` (Windows) or `Cmd + Enter` (Mac).

As discussed in our virtual session, the full code as a pure text document is displayed when you click "Source" button instead of "Visual" in the top left.

------------------------------------------------------------------------

## 2. R Fundamentals: Math, Variables, and Vectors

### Basic Arithmetic

R can function as a highly advanced calculator. Let's run a simple calculation.

```{r}
# Lines starting with a hash (#) are comments. R ignores them.
# They are used to leave notes inside the code chunks.
150 / 3 + 12
```

### The Assignment Operator

To save a value so we can use it later, we assign it to a **variable**. A variable is simply a named storage container. We use the assignment operator `<-` (a less-than sign followed by a hyphen).

Run the chunk below and watch the **Environment** pane (Top Right).

```{r}
# Assign the value 42 to a variable named my_age
my_age <- 42

# Assign text (a "string") to a variable
patient_name <- "John Doe"

# To view what is inside a variable, type its name and run the line
my_age
```

### Vectors

Often, we need to store more than one number at a time. A **vector** is a one-dimensional array (a sequence) of data elements of the same basic type. We use the `c()` function, which stands for "combine" or "concatenate", to create a vector.

```{r}
# Create a vector of multiple patient ages
patient_ages <- c(25, 34, 45, 29, 52)

# Calculate the average (mean) of the vector
mean(patient_ages)
```

------------------------------------------------------------------------

## 3. Managing Your Workspace

As you work, your Environment will fill up with variables. It is crucial to know how to manage this memory.

```{r}
# 1. List all variables currently in memory
ls()

# 2. Remove a specific variable using the rm() function
rm(my_age)

# 3. The Nuclear Option: Clear the entire environment.
# This is highly recommended at the start of a new analysis to prevent old data from interfering.
rm(list = ls())
```

------------------------------------------------------------------------

## 4. Importing and Inspecting Data

In statistics, data is rarely typed out manually. We import it from external files. The most common format is a CSV (Comma-Separated Values) file.

When we import a table into R, it is stored as a **Data Frame**. A data frame is a two-dimensional structure where columns represent variables and rows represent observations.

*Note: Ensure the file `exercise_hr.csv` is saved in the exact same folder as this Quarto document.*

```{r}
# Import the data and assign it to a variable named 'df' (short for data frame)
df <- read.csv("exercise_hr.csv")
```

Before visualizing data, we must inspect it to understand its structure. R provides several functions for this.

```{r}

# View the dimensions (Number of Rows, then Number of Columns)
dim(df)
```

```{r}

# Peek at the first 6 rows to verify column names and data types
head(df)

# Get a statistical summary of every column (Min, Max, Mean, Quartiles)
summary(df)
```

------------------------------------------------------------------------

## 5. Data Visualization

Visualizations allow us to identify patterns, outliers, and relationships. We access specific columns within our data frame using the `$` operator (e.g., `df$ColumnName`).

### A. Histograms (Distribution of a Numeric Variable)

A histogram groups numerical data into bins to show its distribution. Let's analyze the distribution of exercise minutes and resting heart rates.

```{r}

# Histogram 1: Exercise Minutes
hist(df$exercise_min,
     main = "Distribution of Daily Exercise",
     xlab = "Exercise (minutes/day)",
     ylab = "Frequency",
     col = "steelblue",      # Fills the bars with color
     border = "white")       # Changes the border color of the bars
```

```{r}

# Histogram 2: Resting Heart Rate
hist(df$resting_hr,
     main = "Distribution of Resting Heart Rate",
     xlab = "Resting HR (bpm)",
     col = "lightgreen",
     border = "white")
```

### B. Bar Plots (Counts of a Categorical Variable)

A bar plot is used for categorical data (text labels or distinct groups). We cannot plot raw text; we must first count how many observations fall into each category using the `table()` function.

```{r}
# 1. Create a frequency table of the glycemic status
counts <- table(df$gly_status)

# 2. View the counts
counts

# 3. Plot the aggregated data
barplot(counts,
        main = "Patient Glycemic Status",
        xlab = "Status Category",
        ylab = "Number of Patients",
        col = c("coral", "gold", "lightblue")) # Applying multiple colors
```

### C. Boxplots (Comparing Numeric Data Across Categories)

We discussed quartiles during our first virtual session.

A boxplot shows the median, quartiles, and outliers of a numeric variable. It is highly effective for comparing a numeric variable across different categories.

We use the tilde `~` operator, which translates to "explained by" or "grouped by".

```{r}
# Syntax: boxplot(Numeric_Variable ~ Categorical_Variable)
boxplot(df$resting_hr ~ df$gly_status,
        main = "Resting HR by Glycemic Status",
        ylab = "Resting HR (bpm)",
        xlab = "Glycemic Status",
        col = "plum")
```

### D. Scatter Plots (Relationship Between Two Numeric Variables)

A scatter plot displays individual data points on an X and Y axis to reveal correlations.

```{r}
# Syntax: plot(X_Variable, Y_Variable)
plot(df$exercise_min, df$resting_hr,
     main = "Exercise Duration vs. Resting Heart Rate",
     xlab = "Exercise (minutes/day)",
     ylab = "Resting HR (bpm)",
     pch = 19,               # Uses solid circles instead of empty rings
     col = rgb(0, 0, 0, 0.4)) # Uses black (0,0,0) with 40% opacity (0.4) to handle overlapping points
```

------------------------------------------------------------------------

## 6. Rendering Your Report

You have successfully written code, managed memory, imported data, and built visualizations.

To convert this entire interactive document into a finalized, shareable HTML report, click the **Render** button at the top of the Source pane. RStudio will execute the code chronologically and compile the output.
