---
title: "Interactive Distribution Limits"
format: html
execute:
  echo: true
  warning: false
---

## 📦 Computational Dependencies and Setup

This initial cell clears the workspace environment to prevent variable interference, then loads the plotting mechanisms and spatial layout utilities required to construct the charts.

```{r}
#| label: setup

# Clear workspace memory to maintain environment hygiene
rm(list = ls())

library(ggplot2)   # Core grammar of graphics engine
library(patchwork) # Grid assembly layout utility
```

## 📊 Output 1: Standalone Variable Histogram

Modify the parameters inside this chunk to instantly regenerate the sample volume and bin resolution for the single plot.

```{r}
#| label: single-histogram
#| fig-width: 8
#| fig-height: 5

# 🛠️ USER PARAMETERS: Modify these values to regenerate the plot
DRAWS_SINGLE <- 1000000  # Total random observations drawn
BINS_SINGLE  <- 20      # Number of discrete bars across the domain

# Execute simulation
set.seed(42)
data_single <- data.frame(x = rnorm(DRAWS_SINGLE))

# Render visualization
ggplot(data_single, aes(x = x)) +
  # after_stat(density) normalizes the area under the bars to equal 1 (the continuous scale)
  geom_histogram(aes(y = after_stat(density)), bins = BINS_SINGLE, 
                 fill = "steelblue", color = "black", alpha = 0.7) +
  # Overlay the limiting theoretical continuous standard normal curve
  stat_function(fun = dnorm, color = "red", linewidth = 1) +
  labs(title = paste("Single Target Histogram (Draws =", DRAWS_SINGLE, "| Bins =", BINS_SINGLE, ")"),
       x = "Value (x)", y = "Density") +
  theme_minimal()
```

## 🎛️ Output 2: Four-Panel Convergence Matrix

Modify the parameters below to control the uniform draw size and individual bin properties across the four quadrant subplots.

```{r}
#| label: matrix-histogram
#| fig-width: 10
#| fig-height: 8

# 🛠️ USER PARAMETERS: Modify these values to control the matrix grid
DRAWS_MATRIX <- 1000000  # Uniform sample size applied to all subplots
SUBPLOT_BIN_1 <- 5       # Bin count for Top-Left Panel
SUBPLOT_BIN_2 <- 15      # Bin count for Top-Right Panel
SUBPLOT_BIN_3 <- 40      # Bin count for Bottom-Left Panel
SUBPLOT_BIN_4 <- 100     # Bin count for Bottom-Right Panel

# Execute simulation
set.seed(42)
data_matrix <- data.frame(x = rnorm(DRAWS_MATRIX))

# Automation function to build standard subplots systematically
build_panel <- function(df, target_bins, panel_title) {
  ggplot(df, aes(x = x)) +
    geom_histogram(aes(y = after_stat(density)), bins = target_bins, 
                   fill = "seagreen", color = "black", alpha = 0.6, linewidth = 0.2) +
    stat_function(fun = dnorm, color = "red", linewidth = 0.8) +
    labs(title = panel_title, x = "Value (x)", y = "Density") +
    theme_minimal()
}

# Compile individual quadrant layouts
p1 <- build_panel(data_matrix, SUBPLOT_BIN_1, paste("Panel A: Bins =", SUBPLOT_BIN_1))
p2 <- build_panel(data_matrix, SUBPLOT_BIN_2, paste("Panel B: Bins =", SUBPLOT_BIN_2))
p3 <- build_panel(data_matrix, SUBPLOT_BIN_3, paste("Panel C: Bins =", SUBPLOT_BIN_3))
p4 <- build_panel(data_matrix, SUBPLOT_BIN_4, paste("Panel D: Bins =", SUBPLOT_BIN_4))

# Assemble the 2x2 multi-panel layout using patchwork syntax
(p1 | p2) / (p3 | p4) + 
  plot_annotation(title = paste("Asymptotic Convergence Matrix (Total Matrix Draws =", DRAWS_MATRIX, ")"))
```

## 📐 Numerical Integration: Computing Areas in R

The symbolic area represented by $\int_{a}^{b} f(x) \, dx$ can be computed numerically in R using the built-in `integrate()` function. This allows us to determine the precise area under any general continuous curve. The accompanying plot visually demarcates the boundaries of the interval being integrated.

```{r}
#| label: general-area-computation

# Define the target function
target_function <- function(x) { sin(x) }

# Define the lower boundary (a) and upper boundary (b)
lower_limit <- 0
upper_limit <- pi

# Calculate the area under the function between the specified limits
integration_result <- integrate(f = target_function, lower = lower_limit, upper = upper_limit)

# Display the comprehensive output
print(integration_result)

# Define plotting range to provide visual context around the boundaries
plot_min <- lower_limit - 1
plot_max <- upper_limit + 1

# Render visualization of the function with integration limits demarcated
ggplot(data.frame(x = c(plot_min, plot_max)), aes(x = x)) +
  stat_function(fun = target_function, color = "blue", linewidth = 1) +
  geom_vline(xintercept = c(lower_limit, upper_limit), color = "red", linetype = "dashed", linewidth = 0.8) +
  labs(title = "Area Computation: f(x) = sin(x)",
       x = "x", y = "f(x)") +
  theme_minimal()
```

## 📈 Visualizing and Verifying the Normal Distribution

By defining universal integration boundaries, we can construct the normal distribution curve and simultaneously calculate its area, expected value, and variance.

Changing the `LOWER_BND` and `UPPER_BND` parameters will recalculate the integrals for that specific interval and automatically draw vertical boundary lines on the plot to visually demarcate the area being considered.

```{r}
#| label: normal-visualization-and-integration

# Define physical distribution parameters
mu_val <- 5
sigma_val <- 2

# Define integration limits (Default: -Inf to Inf to verify total distribution)
LOWER_BND <- -Inf
UPPER_BND <- Inf

# Define the normal probability density function (PDF)
pdf_func <- function(x) { dnorm(x, mean = mu_val, sd = sigma_val) }

# 1. Area: Integral of f(x)
calc_area <- integrate(pdf_func, lower = LOWER_BND, upper = UPPER_BND)$value
cat("Computed Area:", calc_area, "\n")

# 2. Expected Value: Integral of x * f(x)
expected_func <- function(x) { x * pdf_func(x) }
calc_mean <- integrate(expected_func, lower = LOWER_BND, upper = UPPER_BND)$value
cat("Target Mean:", mu_val, "| Computed Expected Value:", calc_mean, "\n")

# 3. Variance: Integral of (x - mu)^2 * f(x)
variance_func <- function(x) { (x - mu_val)^2 * pdf_func(x) }
calc_var <- integrate(variance_func, lower = LOWER_BND, upper = UPPER_BND)$value
cat("Target Variance:", sigma_val^2, "| Computed Variance:", calc_var, "\n")

# Render the normal distribution curve spanning 4 standard deviations
plot_min <- mu_val - 4 * sigma_val
plot_max <- mu_val + 4 * sigma_val

ggplot(data.frame(x = c(plot_min, plot_max)), aes(x = x)) +
  stat_function(fun = pdf_func, color = "blue", linewidth = 1) +
  # Add vertical lines to demarcate the integration limits (visible if finite)
  geom_vline(xintercept = c(LOWER_BND, UPPER_BND), color = "red", linetype = "dashed", linewidth = 0.8) +
  labs(title = paste("Normal Distribution (mu =", mu_val, ", sigma =", sigma_val, ")"),
       x = "Value (x)", y = "Density f(x)") +
  theme_minimal()
```

## 📉 Computing Probabilities via Z-Scores

For any normal distribution, the probability (area) to the left of a specific threshold depends entirely on the standardized Z-score. Because of this, we do not need to integrate custom functions repeatedly.

In R, the `pnorm(z)` function natively computes the left-tail area of a standard normal distribution based solely on the Z-score. We use this to instantly calculate the probabilities associated with the empirical rule.

```{r}
#| label: empirical-rule-pnorm

# 1 Standard Deviation (Z between -1 and 1)
area_1 <- pnorm(1) - pnorm(-1)
cat("Area within 1 sigma (+/- 1Z):", round(area_1 * 100, 2), "%\n")

# 2 Standard Deviations (Z between -2 and 2)
area_2 <- pnorm(2) - pnorm(-2)
cat("Area within 2 sigma (+/- 2Z):", round(area_2 * 100, 2), "%\n")

# 3 Standard Deviations (Z between -3 and 3)
area_3 <- pnorm(3) - pnorm(-3)
cat("Area within 3 sigma (+/- 3Z):", round(area_3 * 100, 2), "%\n")
```
